Cross Validation Approaches for Penalized Cox Regression

### 2019-05-17 main.tex
Editing real data section (in progress)

%\note{What's with all the \textbackslash par commands?  They don't seem to do anything.}

@@

\par The final metric we included in our study was Harrell's C Index \citep{HarrellJr1984}. The C Index is a rank-based statistic that measures the concordance between the linear predictor of the selected model and the observed outcome. Suppose we arbitrarily select two individuals; if the individual with the higher predicted risk also died first, then this would be considered as a concordant pair. The C Index considers all possible such pairs and estimates of proportion of those pairs that are concordant. A C Index of 0.5 means the prediction is no better than flipping a coin and C Index of 1 means perfect concordance. We computed Harrell's C Index using the independent test set of 1000 individuals described above.

\begin{figure}[ht]
  \centering
  \includegraphics[height= 7cm ]{./manuscript_figure/figure_2_new.png}
  \caption{\label{Fig:mse-brier-c} Simulation comparing 4 CV methods. The horizontal axis in all three plots is $s$, which represents the signal strength in the data. The value of $s$ is varied from 0.4 to 0.9.  $\log(\text{MSE ratio})$, relative to the oracle modle is plotted in the left panel. Out-of-sample Brier scores are plotted in the middle. Out-of-sample Harrell's C index is plotted in the right panel. For each simulated data set, n = 120, p = 1000. Expected censoring percentage is 10$\%$. 
%\note{I really think the x axis should be Signal (s) or Signal strength (s)}
}
\end{figure}

@@

\par Figure~\ref{Fig:mse-brier-c} illustrates the results of this simulation study. In each panel, the horizontal axis is the magnitude of the non-zero $\beta$ coefficients.  For all four metrics, the overall pattern is the same (KL scores not shown in Figure~\ref{Fig:mse-brier-c} due to space constraints, but they were very similar to Brier scores).  Although the four cross-validation methods performed relatively similarly, the standard and linear predictor approaches had the best performance across all metrics: lower MSE, better Brier scores, and higher Harrell's C Index.

@@

\caption{\label{Tab:sim} Simulation comparing four cross-validation methods with varying dimension and signal strength}
\centering
\begin{tabular}[t]{cccccccccc}
\toprule
 n& p& $\#$ Non-Zero & Signal & Method & $\lambda$ & log(MSE Ratio) &Brier Score & Kullback Leibler & Harrell's C Index\\
\midrule
150 & 50 & 5 & weak & V VH  & 0.098 (0.030) & 1.561 (0.851) & 0.245 (0.014) & 0.685 (0.030) & 0.610 (0.035) \\
    &    &   &     & Standard  & 0.088 (0.025) & 1.531 (0.838) & 0.244 (0.013) & 0.683 (0.029) & 0.614 (0.029) \\
    &    &   &     & LinearPred  & 0.092 (0.029) & 1.543 (0.846) & 0.245 (0.014) & 0.684 (0.030) & 0.612 (0.034) \\
    &    &   &     & DevResid  & 0.098 (0.029) & 1.547 (0.848) & 0.245 (0.014) & 0.684 (0.030) & 0.611 (0.035) \\
\addlinespace
150 & 50 & 5 & strong & V VH  & 0.071 (0.010) & 1.586 (0.989) & 0.176 (0.011) & 0.526 (0.027) & 0.750 (0.012) \\
    &    &   &     & Standard  & 0.066 (0.011) & 1.544 (0.971) & 0.176 (0.011) & 0.526 (0.028) & 0.749 (0.013) \\
    &    &   &     & LinearPred  & 0.067 (0.010) & 1.540 (0.981) & 0.176 (0.011) & 0.525 (0.028) & 0.750 (0.012) \\
    &    &   &     & DevResid  & 0.094 (0.011) & 1.995 (0.841) & 0.181 (0.011) & 0.539 (0.026) & 0.751 (0.013) \\
\addlinespace
400 & 10000 & 20 & weak & V VH  & 0.126 (0.019) & 2.958 (0.416) & 0.251 (0.012) & 0.696 (0.026) & 0.589 (0.044) \\
    &    &   &     & Standard  & 0.103 (0.022) & 2.880 (0.437) & 0.244 (0.015) & 0.680 (0.031) & 0.603 (0.039) \\
    &    &   &     & LinearPred  & 0.105 (0.023) & 2.886 (0.437) & 0.244 (0.015) & 0.681 (0.031) & 0.602 (0.039) \\
    &    &   &     & DevResid  & 0.107 (0.022) & 2.895 (0.433) & 0.245 (0.014) & 0.683 (0.03) & 0.602 (0.039) \\
\addlinespace
400 & 10000 & 20 & strong & V VH  & 0.084 (0.012) & 3.799 (0.548) & 0.176 (0.03) & 0.528 (0.069) & 0.763 (0.048) \\
    &    &   &     & Standard  & 0.059 (0.007) & 3.455 (0.578) & 0.154 (0.019) & 0.468 (0.049) & 0.780 (0.031) \\
    &    &   &     & LinearPred  & 0.060 (0.007) & 3.479 (0.574) & 0.155 (0.019) & 0.470 (0.049) & 0.780 (0.030) \\
    &    &   &     & DevResid  & 0.076 (0.018) & 3.708 (0.518) & 0.166 (0.024) & 0.502 (0.057) & 0.772 (0.049) \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}[h]
  \centering
  \includegraphics[height= 15cm ]{./manuscript_figure/figure_auc.png}
  \caption{\label{Fig:cvauc}Comparing Partial Likelihood and Harrell's C Index both built from Cross-validated Linear Predictors. The horizontal axis represents the strength of signal $s$ in the data. The left figures are under the low dimension scenarios where p = 50 and n = 150. The right figures are under the high dimension scenarios where p = 1000 and n = 120. The expected percentages of censoring for both scenarios are 30 $\%$. The top row are box plots of the selected tuning parameter$\lambda$. The mid row plots the $\log(\text{relative Squared Error Loss})$, with respect to the oracle model. The bottom row plots the Brier Scores. }
\end{figure}

@@

For this analysis, we fit elastic net models to the data in addition to lasso models, as these models often offer improvement over the lasso when many correlated genes are involved \citep{Zou2005}. The clinical variables stage, age and gender were included in the model without penalization, as it is reasonable to assume that all three have an effect on survival. Besides the regularization parameter $\lam$, the elastic net model has an extra tuning parameter $\alpha$, where $\alpha=1$ corresponds to lasso and $\alpha=0$ corresponds to ridge regression. For each $\alpha = 0.3, 0.5, 0.7, 0.9 \text{ and } 1$, we selected a regularization parameter $\lam$ that has the lowest cross-validation error for each of the four cross-validation methods described in Section~\ref{Sec:methods}. We reported the cross-validated C-Index and Brier Scores of thoses models in Table~\ref{tb:shedden}. With the highest C Index and lowest Brier Score, the model selected by the linear predictor approach at $\alpha = 0.3$ has best predictive performance among all of them.

@@

For each fixed $\alpha$, we compared the cross-validated errors(CVE) from the four cross-validation methods along the solution path. At $\alpha = 0.3$, CVEs are rescaled and plotted for all four met  hods at the same spectrum of $\lambda$ values In Figure~\ref{fig:CVE}. A $\lambda$ is selected when a CVE curve reaches its lower point. The blue line and the green line, which represent the CVE curves of linear predictor approach and standard approah, have more curvature near theirs lowest point, making it easier to pick out the minimum points. The selected $\lambda$ values, represented by the vertical dashed lines, and their corresponding models, reflected what is shown in simulation studies. The linear predictor approach, selecting smallest $\lambda$ values, was most liberal and selected 85 genes. While the standard approach and deviance residual approach selected 68 genes and 60 genes respectively, Verweij and Van Houwelingen's approach only selected 34 genes and performed most conservatively on this data set. Even though we only illustrated the models at $\alpha = 0.3$, same patterns are repeated when $\alpha$ is fixed at other values.

>>>>

@@

\par The final metric we included in our study was Harrell's C Index \citep{HarrellJr1984}. The C Index is a rank-based statistic that measures the concordance between the linear predictor of the selected model and the observed outcome. Suppose we arbitrarily select two individuals; if the individual with the higher predicted risk also died first, then this would be considered as a concordant pair. The C Index considers all possible such pairs and estimates of proportion of those pairs that are concordant. A C Index of 0.5 means the prediction is no better than flipping a coin and C Index of 1 means perfect concordance. We computed the C Index using the independent test set of 1000 individuals described above.

\begin{figure}[ht]
  \centering
  \includegraphics[height= 7cm ]{./manuscript_figure/figure_2_new.png}
  \caption{\label{Fig:mse-brier-c} Simulation comparing 4 CV methods. The horizontal axis in all three plots is $s$, which represents the signal strength in the data. The value of $s$ is varied from 0.4 to 0.9.  $\log(\text{MSE ratio})$, relative to the oracle modle is plotted in the left panel. Out-of-sample Brier scores are plotted in the middle. Out-of-sample C index is plotted in the right panel. For each simulated data set, n = 120, p = 1000. Expected censoring percentage is 10$\%$. 
%\note{I really think the x axis should be Signal (s) or Signal strength (s)}
}
\end{figure}

@@

\par Figure~\ref{Fig:mse-brier-c} illustrates the results of this simulation study. In each panel, the horizontal axis is the magnitude of the non-zero $\beta$ coefficients.  For all four metrics, the overall pattern is the same (KL scores not shown in Figure~\ref{Fig:mse-brier-c} due to space constraints, but they were very similar to Brier scores).  Although the four cross-validation methods performed relatively similarly, the standard and linear predictor approaches had the best performance across all metrics: lower MSE, better Brier scores, and higher C Index.

@@

\caption{\label{Tab:sim} Simulation comparing four cross-validation methods with varying dimension and signal strength}
\centering
\begin{tabular}[t]{cccccccccc}
\toprule
 n& p& $\#$ Non-Zero & Signal & Method & $\lambda$ & log(MSE Ratio) &Brier Score & Kullback Leibler & C Index\\
\midrule
150 & 50 & 5 & weak & V VH  & 0.098 (0.030) & 1.561 (0.851) & 0.245 (0.014) & 0.685 (0.030) & 0.610 (0.035) \\
    &    &   &     & Standard  & 0.088 (0.025) & 1.531 (0.838) & 0.244 (0.013) & 0.683 (0.029) & 0.614 (0.029) \\
    &    &   &     & LinearPred  & 0.092 (0.029) & 1.543 (0.846) & 0.245 (0.014) & 0.684 (0.030) & 0.612 (0.034) \\
    &    &   &     & DevResid  & 0.098 (0.029) & 1.547 (0.848) & 0.245 (0.014) & 0.684 (0.030) & 0.611 (0.035) \\
\addlinespace
150 & 50 & 5 & strong & V VH  & 0.071 (0.010) & 1.586 (0.989) & 0.176 (0.011) & 0.526 (0.027) & 0.750 (0.012) \\
    &    &   &     & Standard  & 0.066 (0.011) & 1.544 (0.971) & 0.176 (0.011) & 0.526 (0.028) & 0.749 (0.013) \\
    &    &   &     & LinearPred  & 0.067 (0.010) & 1.540 (0.981) & 0.176 (0.011) & 0.525 (0.028) & 0.750 (0.012) \\
    &    &   &     & DevResid  & 0.094 (0.011) & 1.995 (0.841) & 0.181 (0.011) & 0.539 (0.026) & 0.751 (0.013) \\
\addlinespace
400 & 10000 & 20 & weak & V VH  & 0.126 (0.019) & 2.958 (0.416) & 0.251 (0.012) & 0.696 (0.026) & 0.589 (0.044) \\
    &    &   &     & Standard  & 0.103 (0.022) & 2.880 (0.437) & 0.244 (0.015) & 0.680 (0.031) & 0.603 (0.039) \\
    &    &   &     & LinearPred  & 0.105 (0.023) & 2.886 (0.437) & 0.244 (0.015) & 0.681 (0.031) & 0.602 (0.039) \\
    &    &   &     & DevResid  & 0.107 (0.022) & 2.895 (0.433) & 0.245 (0.014) & 0.683 (0.03) & 0.602 (0.039) \\
\addlinespace
400 & 10000 & 20 & strong & V VH  & 0.084 (0.012) & 3.799 (0.548) & 0.176 (0.03) & 0.528 (0.069) & 0.763 (0.048) \\
    &    &   &     & Standard  & 0.059 (0.007) & 3.455 (0.578) & 0.154 (0.019) & 0.468 (0.049) & 0.780 (0.031) \\
    &    &   &     & LinearPred  & 0.060 (0.007) & 3.479 (0.574) & 0.155 (0.019) & 0.470 (0.049) & 0.780 (0.030) \\
    &    &   &     & DevResid  & 0.076 (0.018) & 3.708 (0.518) & 0.166 (0.024) & 0.502 (0.057) & 0.772 (0.049) \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}[h]
  \centering
  \includegraphics[height= 15cm ]{./manuscript_figure/figure_auc.png}
  \caption{\label{Fig:cvauc}Comparing Partial Likelihood and C Index both built from Cross-validated Linear Predictors. The horizontal axis represents the strength of signal $s$ in the data. The left figures are under the low dimension scenarios where p = 50 and n = 150. The right figures are under the high dimension scenarios where p = 1000 and n = 120. The expected percentages of censoring for both scenarios are 30 $\%$. The top row are box plots of the selected tuning parameter$\lambda$. The mid row plots the $\log(\text{relative Squared Error Loss})$, with respect to the oracle model. The bottom row plots the Brier Scores. }
\end{figure}

@@

For this analysis, we fit elastic net models to the data in addition to lasso models, as these models often improve upon the lasso when many correlated genes are involved \citep{Zou2005}. The clinical variables stage, age and gender were included in the model without penalization, as it is reasonable to assume that all three have an effect on survival. In addition to the regularization parameter $\lam$, the elastic net model has an extra tuning parameter $\alpha$, where $\alpha=1$ corresponds to lasso and $\alpha=0$ corresponds to ridge regression. For $\alpha \in \{0.3, 0.5, 0.7, 0.9, 1\}$, we selected a regularization parameter $\lam$ that has the lowest cross-validation error for each of the four cross-validation methods described in Section~\ref{Sec:methods}.  Since the cross-validation metrics for each method are on different scales and thus not directly comparable, Table~\ref{tb:shedden} reports the cross-validated C-Index and Brier Scores for the CV-selected model at each value of $\alpha$.  As shown in the table, the elastic net model with $\alpha=0.3$ and $\lam$ selected by the cross-validated linear predictor approach has the best performance in terms of C Index and Brier Score.

@@

For each fixed $\alpha$, we compared the cross-validation errors (CVE) from the four cross-validation methods along the solution path. At $\alpha = 0.3$, CVEs are rescaled and plotted for all four methods along the same grid of $\lambda$ values in Figure~\ref{fig:CVE}. The selected $\lambda$ values, where a given CVE curve reaches its lowest point, are indicated with dotted lines.  The CVE curves for the standard and linear predictor approaches have greater curvature near their minima, and thus offer more clear support for the selected value of $\lam$.  In comparison, the deviance residual and V\&VH approaches are flatter near their minima, supporting a considerably wider range of $\lam$ values and leading to greater model uncertainty.   The selected $\lambda$ values, represented by the vertical dashed lines, and their corresponding models, reflected what is shown in simulation studies. The linear predictor approach, selecting smallest $\lambda$ values, was most liberal and selected 85 genes. While the standard approach and deviance residual approach selected 68 genes and 60 genes respectively, Verweij and Van Houwelingen's approach only selected 34 genes and performed most conservatively on this data set. Even though we only illustrated the models at $\alpha = 0.3$, same patterns are repeated when $\alpha$ is fixed at other values.
